In the Crucible: Massive Compact Galaxies with High-Velocity Outflows and Eddington-Limited Star Formation

### 2012-04-19 trunk/tex/sigmasfr/ms.tex
working on discussion text

Based on infrared photometry from the Wide-field Infrared Survey
Explorer (WISE) and optical morphologies from the Hubble Space
Telescope (HST), we present the discovery of compact, obscured star
formation in galaxies that exhibit $\gtrsim1000$~\kms\ outflows.  The
implied star formation rate (SFR) surface densities approach
$\sigmasfr\approx5000$~\msun~yr$^{-1}$~kpc$^{-2}$, comparable to the
Eddington limit for a starburst where radiation pressure on dust
grains overcomes gravity.  We argue that feedback associated with this
starburst event, in the form of radiation pressure and supernova ram
pressure, is a likely mechanism for launching the high-velocity
outflows.

@@

\keywords{keywords}

@@

Galactic winds are ubiquitous in star-forming galaxies at all
redshifts and generally exhibit outflow velocities in the
100--500~\kms\ range
\citep[e.g.,][]{hec00,sha03,mar05,rup05,wei09,rub10,ste10}.  Thus, the
discovery of $|v|>1000$~\kms\ outflows in a sample of massive
($\textnormal{\mstar}\approx10^{11}$~\msun) post-starburst galaxies at
$z\sim0.6$ by \citet{tre07} suggested that a more energetic source,
such as feedback from an accreting supermassive black hole
\citep[e.g.,][]{sil98,dim05,deb12}, could have been responsible for
launching the winds.

@@

The compact sizes ($r_e\approx100$~pc) and large star formation rates
($\textnormal{SFR}\approx500$~\msun) for the galaxies described above
implies extremely large star-formation rate surface densities
($\sigmasfr\approx5000$~\units).  To place these galaxies in context,
we plot their $\sigmasfr$ values as a function of stellar mass in
Figure~\ref{fig:sigmasfr}.  We include comparison samples of
$\sim10^5$ star-forming galaxies at $0.5<z<1.5$ from \citet{wuy11} and
a few dozen gas-rich mergers at $z<0.3$ including 32 ULIRGs from
\citet{vei06}, five Lyman break analogs with dominant central objects
from \citet{ove09}, and the local compact starburst Arp 220
\citep{sco97, ken98, rod08}.  We also mark the empirical threshold for
launching winds \citep[$\sigmasfr\approx0.1$~\units,][]{hec02}, the
90th percentile limit for the surface brightness of starbursts
measured using UV, H$\alpha$, far-IR, and radio continuum emission
\citep[$\sigmasfr\approx45$~\units][]{meu97}, and the theoretical
limit for a starburst limited by feedback from radiation pressure
\citep[$\sigmasfr\approx2000$~\units,][]{mur05,tho05,hop10}.  The most
luminous, compact starbursts in our sample exhibit SFR surface
densities that reach the Eddington limit, suggesting that their growth
is being regulated by momentum input from massive stars.

@@

They are rare and thus represent an unusual or short-lived phase.  Gas
consumption time for starbursts.  Connection to ellipticals.
Conceivable that they could grown in size sufficiently by $z=0$?

@@

Connection to models.  How such large SFRs given the powerful wind?

@@

@@

>>>>

We present the discovery of compact, obscured star formation in
galaxies that exhibit $\gtrsim1000$~\kms\ outflows.  Using optical
morphologies from the Hubble Space Telescope and infrared photometry
from the Wide-field Infrared Survey Explorer, we estimate star
formation rate (SFR) surface densities that approach
$\sigmasfr\approx5000$~\msun~yr$^{-1}$~kpc$^{-2}$.  This is comparable
to the Eddington limit from radiation pressure on dust grains.  We
argue that feedback associated with a compact starburst in the form of
radiation pressure from massive stars and ram pressure from supernovae
and stellar winds is sufficient to launch the high-velocity outflows
we observe, without the need to invoke feedback from an active
galactic nucleus.

@@

\keywords{galaxies: evolution --- galaxies: kinematics and dynamics
  --- galaxies: ISM --- galaxies: starburst}

@@

Galactic winds are ubiquitous in star-forming galaxies at all
redshifts and generally exhibit outflow velocities in the
100--500~\kms\ range
\citep[e.g.,][]{hec00,sha03,mar05,rup05,wei09,rub10,ste10}.  Thus, the
discovery of $|v|>1000$~\kms\ outflows in a sample of massive
($\textnormal{\mstar}\approx10^{11}$~\msun) post-starburst galaxies at
$z\sim0.6$ by \citet{tre07} suggested that a more energetic source,
such as feedback from an accreting supermassive black hole
\citep[e.g.,][]{sil98,dim05,deb12}, could have been responsible for
launching the winds \citep[see][for a recent review]{fab12}.

@@

The compact sizes ($r_e\approx100$~pc) and large star formation rates
($\textnormal{SFR}\approx300$~\msun) for the galaxies described above
implies extremely large star-formation rate surface densities
($\sigmasfr\approx5000$~\units).  To place these galaxies in context,
we plot their $\sigmasfr$ values as a function of stellar mass in
Figure~\ref{fig:sigmasfr}.  We include comparison samples of
$\sim10^5$ star-forming galaxies at $0.5<z<1.5$ from \citet{wuy11} and
a few dozen gas-rich mergers at $z<0.3$ including 32 ULIRGs from
\citet{vei06}, five Lyman break analogs with dominant central objects
from \citet{ove09}, and the local compact starburst Arp 220
\citep{sco97, ken98, rod08}.  We also mark the empirical threshold for
launching winds \citep[$\sigmasfr\approx0.1$~\units,][]{hec02}, the
90th percentile limit for the surface brightness of starbursts
measured using UV, H$\alpha$, far-IR, and radio continuum emission
\citep[$\sigmasfr\approx45$~\units][]{meu97}, and the theoretical
limit for a starburst limited by feedback from radiation pressure
\citep[$\sigmasfr\approx2000$~\units,][]{mur05,tho05,hop10}.  The most
luminous, compact starbursts in our sample exhibit SFR surface
densities that reach the Eddington limit, suggesting that their growth
is being regulated by momentum input from massive stars.

@@

The high-$\sigmasfr$ galaxies in our sample constitute a rare
population (see Figure~\ref{fig:sigmasfr}), suggesting that they
represent an unusual or short-lived phase.  Mergers of gas-rich
galaxies are a viable mechanism for producing compact starbursts
(e.g., Mihos+94), and such gas-rich major mergers are rare at
$z\sim0.6$ due to the decline in both the merger rate and the gas
fraction of galaxies since $z\sim1$.  The high-$\sigmasfr$ galaxies in
particular are in a phase where there is strong feedback but star
formation has not been quenched.  The length of this phase may be set
by the gas consumption timescale or the timescale for feedback to
suppress subsequent star formation.  Based on the Kennicutt--Schmidt
(K-S) relation \citep{ken98}, a galaxy with $r_e\approx100$~pc,
$\textnormal{SFR}\approx300$~\msun, and
$\sigmasfr=5000$~\units\ (assuming half of the star formation occurs
within the effective radius) would have a gas surface density of
$\Sigma_{gas}\sim2\times10^5$~\msun~pc$^{-2}$, corresponding to
$M_{gas}\sim5\times10^{9}$~\msun\ inside 100~pc, which would be
consumed on a timescale $\sim30$~Myr.  It would be worthwhile to
directly measure the molecular gas masses and kinematics of these
extreme sources with CO observations to test this scenario.

The vertical sequence in Figure~\ref{fig:sigmasfr} may be related to
an evolutionary sequence, with the smaller $\sigmasfr$ galaxies being
further along in their evolution towards a post-starburst phase.
However, it is interesting to note that all 12/29 galaxies above the
$\sigmasfr\approx45$~\units\ \citet{meu97} limit exhibit outflows
(with median centroid velocity $v=-1500$~\kms), while all 7/29
galaxies without detected outflows have $\sigmasfr<30$~\units.  If a
compact starburst is a requirement for the production of high-velocity
outflows, it may be that these latter sources have not gone through
such a phase (see A. Robaina, et al, in preparation for a discussion
of the relationship between outflow velocity, galaxy morphology, and
stellar population age).

% although simulations rarely produce starbursts that
% are this compact.

% While such major mergers are rare at $z\sim0.6$ \citep[e.g., the
%   fraction of $M_{*}>3\times10^{10}$~\msun galaxies in $r<30$~kpc
%   pairs is $\sim3$\%][]{bel06}, \citet{rob10} argued that such mergers
% are sufficient to explain the number density evolution of massive red
% sequence galaxies since $z\sim1$ \citep[e.g.,][]{fab07,bro07}.

It is also worhwhile to consider the implications of these galaxies
for models of massive galaxy formation.  Models do not produce
starbursts this small.  Models that produce ellipticals have to invoke
AGN feedback.  Models for size evolution would allow these galaxies to
reach the size--mass relation by $z=0$.

% A merger of two massive, gas-rich galaxies is a viable mechanism for
% producing such a compact starburst (e.g., Mihos+94), but the
% subsequent evolution depends on the assumed feedback prescriptions in
% the model and the gas content of the progenitor galaxies.  It is
% difficult for mergers to produce starbursts that are this compact.
% Models can quench star formation by invoking AGN feedback,

@@

% Connection to models.  How such large SFRs given the powerful wind?

@@

\bibitem[Fabian(2012)]{fab12} Fabian, A.~C.\ 2012,
  arXiv:1204.4114

@@

\bibitem[Robaina et al.(2010)]{rob10} Robaina, A.~R., Bell, E.~F., van
  der Wel, A., et al.\ 2010, \apj, 719, 844
